---
title: Data transformation
---

> 해당 교재는 https://r4ds.hadley.nz/ 에서 확인하실 수 있습니다. 이 문서에 존재하는 모든 인용은 @HadleyWickham2023 입니다.

이 장의 목표는 데이터 프레임을 변환하기 위한 모든 주요 도구인 dplyr 패키지를 사용한 데이터 변환을 소개하는 이 모든 방법(그리고 그 이상!)을 배우게 될 것입니다.

```{r}
library(tidyverse)
library(nycflights13)
library(ggthemes)
```

> [...] which will introduce you to data transformation using the dplyr package and a new dataset on flights that departed from New York City in 2013.

```{r}
flights %>% 
  glimpse()
```

문제를 풀려면 보통 여러 연산자를 결합해야 하는데, 파이프 `|>`를 활용하면, 여러 연산자를 결합해서 사용할 수 있습니다. 파이프는 왼쪽에 있는 것을 오른쪽의 함수에 전달하여 `x |> f(y)`는 `f(x, y)`와 같고, `x |> f(y) |> g(z)`는 `g(f(x, y), z)`와 같습니다. 파이프를 가장 쉽게 발음하는 방법은 "then"입니다. 이렇게 하면 아직 세부 사항을 배우지 않았더라도 다음 코드의 의미를 파악할 수 있습니다.

```{r}
flights |>
  filter(dest == "IAH") |> 
  group_by(year, month, day) |> 
  summarize(
    arr_delay = mean(arr_delay, na.rm = TRUE)
  ) 
```

## Row

### filter

> The most important verbs that operate on rows of a dataset are filter() [...]

```{r}
flights |> 
  filter(dep_delay > 120)
```

```{r}
flights |> 
  filter(month == 1 & day == 1)
flights |> 
  filter(month == 1 | month == 2)
flights |> 
  filter(month %in% c(1, 2))
```

```{r}
jan1 <- flights |> 
  filter(month == 1 & day == 1)
```

### arrange

> arrange() changes the order of the rows based on the value of the columns. [...]

```{r}
flights |> 
  arrange(year, month, day, dep_time)
flights |> 
  arrange(desc(dep_delay))
```

### distinct

> distinct() finds all the unique rows in a dataset, so in a technical sense...

```{r}
flights |> 
  distinct()
flights |> 
  distinct(origin, dest)
```

> Alternatively, if you want to the keep other columns when filtering for unique rows, you can use the .keep_all = TRUE option.

```{r}
flights |> 
  distinct(origin, dest, .keep_all = TRUE)
```

```{r}
flights |>
  count(origin, dest, sort = TRUE)
```

### 연습문제 3.2.5

#### 1

```{r}
# Had an arrival delay of two or more hours
flights |>
  filter(arr_delay >= 120) |>
  arrange(desc(arr_delay))

# Flew to Houston (IAH or HOU)
flights |>
  filter(dest %in% c("IAH", "HOU"))
  
# Were operated by United, American, or Delta
flights |>
  filter(carrier %in% c("UA", "AA", "DL"))

# Departed in summer (July, August, and September)
flights |>
  filter(month %in% c(7, 8, 9))

# Arrived more than two hours late, but didn’t leave late
flights |> 
  filter(arr_delay >= 120 & dep_delay <= 0) |> view()

# Were delayed by at least an hour, but made up over 30 minutes in flight
flights |> 
  filter(dep_delay >= 60 & dep_delay - arr_delay > 30)
```

#### 2

```{r}
flights |> 
  arrange(desc(dep_delay)) |> 
  arrange(sched_dep_time) |>
  relocate(dep_delay, sched_dep_time)
```
#### 3

```{r}
flights |> 
  mutate(speed = distance / (air_time / 60)) |>
  arrange(desc(speed)) |>
  relocate(speed)
```

#### 4

```{r}
flights |> 
  distinct(year, month, day) |>
  nrow()
```
#### 5

```{r}
flights |> 
  arrange(desc(distance)) |>
  relocate(distance)
```
## Columns

### mutate

> The job of mutate() is to add new columns that are calculated from the existing columns. 

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60
  )
```

> [...] We can use the .before argument to instead add the variables to the left hand side.

```{r}
flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .before = 1
  )

flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    speed = distance / air_time * 60,
    .after = day
  )

flights |> 
  mutate(
    gain = dep_delay - arr_delay,
    hours = air_time / 60,
    gain_per_hour = gain / hours,
    .keep = "used"
  )
```

### select

> [...] select() allows you to rapidly zoom in on a useful subset using operations based on the names of the variables:

```{r}
flights |> 
  select(year, month, day)
flights |> 
  select(year:day)
flights |> 
  select(!year:day)
flights |> 
  select(where(is.character))
flights |> 
  select(tail_num = tailnum)
```

### rename

> If you have a bunch of inconsistently named columns and it would be painful to fix them all by hand, check out janitor::clean_names() which provides some useful automated cleaning.

```{r}
flights |> 
  rename(tail_num = tailnum)
```

### relocate

```{r}
flights |> 
  relocate(time_hour, air_time)
flights |> 
  relocate(year:dep_time, .after = time_hour)
flights |> 
  relocate(starts_with("arr"), .before = dep_time)
```
### 연습문제 3.3.5

#### 1

```{r}
flights |> 
  relocate(dep_time, sched_dep_time, dep_delay)
```

#### 2

```{r}
flights |> 
  select(dep_time, dep_delay, arr_time, arr_delay)
flights |> 
  select(starts_with("dep"), starts_with("arr"))
flights |>
  select(dep_time:arr_delay, -contains("sched"))
```

#### 3

```{r}
flights |> 
  select(dep_time, dep_time)
```

#### 4

```{r}
variables <- c("year", "month", "day", "dep_delay", "arr_delay")

flights |> 
  select(any_of(variables))
```

#### 5

```{r}
flights |> 
  select(contains("TIME"))
flights |> 
  select(contains("TIME", ignore.case = FALSE))
```

#### 6

```{r}
flights |>
  rename(air_time_min = air_time) |>
  relocate(air_time_min)
```

## The Pipe

> While both forms have their time and place, the pipe generally produces data analysis code that is easier to write and read.

```{r}
flights |> 
  filter(dest == "IAH") |> 
  mutate(speed = distance / air_time * 60) |> 
  select(year:day, dep_time, carrier, flight, speed) |> 
  arrange(desc(speed))
```

## Groups

> So far you’ve learned about functions that work with rows and columns. dplyr gets even more powerful when you add in the ability to work with groups.

### group_by

> Use group_by() to divide your dataset into groups meaningful for your analysis

```{r}
flights |> 
  group_by(month)
```

### summarize

```{r}
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay)
  )
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE)
  )
flights |> 
  group_by(month) |> 
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE), 
    n = n()
  )
```

### slice_

```{r}
flights |> 
  group_by(dest) |> 
  slice_max(arr_delay, n = 1) |>
  relocate(dest)
```

### Grouping by multiple variables

```{r}
daily <- flights |>  
  group_by(year, month, day)
```

```{r}
daily_flights <- daily |> 
  summarize(n = n())

daily_flights <- daily |> 
  summarize(
    n = n(), 
    .groups = "drop_last"
  )
```

### Ungrouping

```{r}
daily |> 
  ungroup()
daily |> 
  ungroup() |>
  summarize(
    avg_delay = mean(dep_delay, na.rm = TRUE), 
    flights = n()
  )
```


### .by

```{r}
flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n(),
    .by = month
  )

flights |> 
  summarize(
    delay = mean(dep_delay, na.rm = TRUE), 
    n = n(),
    .by = c(origin, dest)
  )
```

### 연습문제 3.5.7

#### 1

```{r}
flights |>
  group_by(carrier) |>
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) |>
  arrange(desc(avg_dep_delay))
```

#### 2

```{r}
flights |> 
  group_by(dest) |> 
  arrange(dest, desc(dep_delay)) |>
  slice_head(n = 5) |>
  relocate(dest, dep_delay)
```

#### 3

```{r}
flights |>
  group_by(hour) |>
  summarize(avg_dep_delay = mean(dep_delay, na.rm = TRUE)) |>
  ggplot(aes(x = hour, y = avg_dep_delay)) + 
  geom_smooth()
```

#### 4

```{r}
flights |> 
  slice_min(dep_delay, n = -5) |>
  relocate(dep_delay)

flights |> 
  slice_min(dep_delay, n = 5) |>
  relocate(dep_delay)

flights |> 
  slice_max(dep_delay, n = -5) |>
  relocate(dep_delay)

flights |> 
  slice_max(dep_delay, n = 5) |>
  relocate(dep_delay)
```

#### 6

```{r}
df <- tibble(
  x = 1:5,
  y = c("a", "b", "a", "a", "b"),
  z = c("K", "K", "L", "L", "K")
)
df |>
  group_by(y)
df |>
  arrange(y)
df |>
  group_by(y) |>
  summarize(mean_x = mean(x))
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x), .groups = "drop")
df |>
  group_by(y, z) |>
  summarize(mean_x = mean(x))
df |>
  group_by(y, z) |>
  mutate(mean_x = mean(x))
```
